The missing puzzle piece: contextual insights for enhanced pharmaceutical supply chain forecasting

### 2025-08-04 manuscript/main.tex
3RD REVISION

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting},
  pdfauthor={Arebu Issa Bilal; Bahman Rostami-Tabar; Harsha Halgamuwe Hewage; Umit Sezer Bititci; Teferi Gedif Fenta},
  pdfkeywords={Forecasting, Pharmaceutical supply chain, Domain
knowlege, Forecast accuracy, Developing countries},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The missing puzzle piece: Contextual insights for enhanced
pharmaceutical supply chain forecasting}
\author[1]{Arebu Issa Bilal%
\corref{cor1}%
}
 \ead{arebu.issa@aau.edu.et} 
\author[2]{Bahman Rostami-Tabar%
%
}
 \ead{rostami-tabarb@cardiff.ac.uk} 
\author[2]{Harsha Halgamuwe Hewage%
%
}
 \ead{HalgamuweHewageHR@cardiff.ac.uk} 
\author[3]{Umit Sezer Bititci%
%
}
 \ead{U.S.Bititci@hw.ac.uk} 
\author[1]{Teferi Gedif Fenta%
%
}
 \ead{tgedif@gmail.com}

\affiliation[1]{organization={Addis Ababa University, Department of
Pharmaceutics and Social Pharmacy, School of
Pharmacy},country={Ethiopia.},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Cardiff University, Data Lab for Social
Good Reserach Group, Cardiff Business
School},country={UK.},countrysep={,},postcodesep={}}
\affiliation[3]{organization={Heriot-Watt University, Edinburgh Business
School, Heriot-Watt
University},country={UK.},countrysep={,},postcodesep={}}

@@

\begin{abstract}
Accurate forecasting in pharmaceutical supply chains is critical for
ensuring the continuous availability of essential medicines,
particularly in developing countries where resource constraints and
logistical challenges are more pronounced. Effective demand forecasting
supports procurement and inventory management, which in turn could help
to prevent stockouts, reduce wastage due to overstocking, and enhance
overall healthcare delivery. In such settings, reliable forecasts, that
acknowledge uncertainties, are essential to ensure that limited
resources are used efficiently to meet health needs. Despite
advancements in forecasting, significant gaps remain in effective
forecasting in poor resource countries. Most existing literature on
pharmaceutical supply chain forecasting focuses primarily on point
estimation of consumption while neglecting the inherent uncertainty of
these forecasts. Additionally, these studies often rely solely on
historical consumption data without considering the broader context that
influences consumption. Furthermore, many works lack rigorous
methodological design, transparency, and reproducibility. This study
addresses these gaps by integrating domain-specific knowledge, gathered
through expert interviews and engagement with key members of the
Ethiopian Pharmaceutical Supply Service (EPSS), into forecasting models.
Using a dataset spanning five years (December 2017 to July 2022) from
EPSS, we developed forecasting models that integrate expert-identified
variables such as stock replenishment schedules, fiscal inventory
counts, and disease outbreaks. Evaluation metrics including Mean
Absolute Scaled Error (MASE), Root Mean Squared Scaled Error (RMSSE),
and Continuous Ranked Probability Score (CRPS) are used to report
forecast accuracy. The findings underscore the significance of
contextual data in developing robust forecasting models that are
adaptable to complex, real-world conditions. Our results also highlight
the effectiveness of foundational time series models for forecasting.
These models are particularly appealing for resource-constrained
countries that may lack advanced analytical expertise. To promote
usability, generalizability, and reproducibility, we share the complete
dataset and code in R and Python, and the entire paper is written in
Quarto via a GitHub repository to facilitate these practices.
\end{abstract}

@@

Demand forecasting in pharmaceutical supply chains is a highly complex
process, shaped by several factors. These include the quality of
available data, the specific position within the supply chain, and
external influences such as market trends, regulatory changes, and
disruptive events such as pandemics, conflicts, flooding
\citep{schneider2010pharmaceutical, hyndman2018forecasting}. Recently,
major donors such as Bill \& Melinda Gates Foundation have invested in
data collection technologies in sub-Saharan Africa, recognizing the
potential of data to provide valuable insights. This investment is
essential and offers significant benefits for improving supply chain
delivery. However, much of the data being collected focuses on
transactional data, such as consumption and sale. Almost all platforms
collecting data tend to overlook the critical contextual factors and
events that influence these transactions---factors that are just as
crucial for accurate forecasting. This presents a major barrier, as
understanding the broader context, including administration procedures,
local policies, disruptions, conflicts, and so on is vital for building
reliable forecasting models. This gap---known as business context and
domain knowledge---requires urgent attention in Pharmaceutical supply
chain forecasting. Without incorporating this knowledge, any forecasting
model is unlikely to produce reliable results. We must prioritize
collecting and integrating this contextual information to improve the
accuracy of demand forecasts in pharmaceutical supply chains. Combining
statistical models with domain insights offers a more refined
forecasting process, as evidenced by research that highlights the value
of judgmental adjustments to enhance forecast accuracy
\citep{fildes2007against, taylor2022artificial, soyiri2013overview}.
Domain knowledge enables analysts to incorporate relevant events, such
as supply chain disruptions, into the forecasting model. This
integration is particularly beneficial in fields like healthcare, where
expert insights can align algorithms with real-world applications
\citep{dash2022review}.

@@

Accurate sales forecasting is crucial in the pharmaceutical industry,
where it directly impacts profit maximization, cost minimization, and
the ability to respond to market changes. Forecasting in healthcare not
only influences clinical decisions but also plays a pivotal role in
supply chain management, ensuring that the right drugs are available
when needed. The challenge lies in balancing customer demand with
inventory costs, which is particularly complex in pharmaceuticals due to
factors like short shelf life and quality constraints
\citep{gupta2000mid, makridakis2020forecasting}.

In this review, a range of forecasting methods has been applied, from
traditional statistical models to advanced machine learning techniques,
with many studies using a combination of simple and advanced methods to
enhance accuracy
\citep{nikolopoulos2016forecasting, zhu2021demand, anusha2014demand}.
Some studies, such as those by \citet{newberne2006holt}, and
\citet{restyana2021analysis}, applied traditional forecasting methods
like the Holt-Winters method, Single Moving Average (SMA), and Single
Exponential Smoothing (SES), which are simpler but effective for certain
applications. In contrast, other researchers explored more advanced
approaches, including hybrid models and machine learning techniques, to
address complex forecasting challenges, as seen in the studies by
\citet{siddiqui2022hybrid}, \citet{de2021lead}, \citet{kim2015demand},
\citet{candan2014demand}, and \citet{ribeiro2017improving}.

Early research in pharmaceutical forecasting primarily focused on
traditional statistical methods. These methods, including ARIMA,
exponential smoothing, and moving averages, were widely used due to
their simplicity and effectiveness in relatively stable environments
\citep{zahra2019forecasting}. These models typically assess forecast
accuracy using metrics like Mean Absolute Percentage Error (MAPE), Mean
Squared Error (MSE), and Root Mean Squared Error (RMSE). For example,
\citet{newberne2006holt} applied the Holt-Winters method to forecast
healthcare data, focusing on prescription trends. Their findings
highlighted the method's utility in short-term planning and resource
management, validated through metrics like MAPE, MAD, and MSD.
Similarly, \citet{restyana2021analysis} compared Single Moving Average
(SMA) and Single Exponential Smoothing (SES) in drug demand forecasting,
concluding that SES provided more accurate results, as indicated by
lower MAD and MSE values.

Recognizing the limitations of traditional models, particularly in
capturing the complexities of pharmaceutical demand (e.g., seasonality,
external influences), researchers began developing hybrid models. These
models combine linear methods like ARIMA with nonlinear techniques such
as neural networks, aiming to improve forecasting accuracy by addressing
diverse data patterns. \citet{khalil2014intelligent} introduced a hybrid
model that integrated ARIMA with neural networks. Their study
demonstrated that this approach significantly outperformed traditional
models, especially in scenarios with limited historical data. The hybrid
model's effectiveness was confirmed by improvements in key performance
indicators like MAPE and RMSE.

The pharmaceutical industry's increasing complexity and the availability
of large datasets have led to the adoption of advanced machine learning
techniques. These methods, such as Support Vector Regression (SVR),
Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer
greater flexibility and predictive power. \citet{van2021using} showed
that incorporating downstream information into machine learning models,
like LASSO and SVR, greatly enhances forecast accuracy. Their study on
multi-echelon supply chains revealed that advanced models, particularly
those integrating external variables, consistently produced lower
forecast errors, as measured by AvgRelRMSE. Similarly,
\citet{rathipriya2023demand} compared shallow neural networks and deep
learning models for drug demand forecasting. The study found that
shallow models outperformed deep learning models for most drug
categories, while ARIMA was more effective for the remaining categories.
This suggests that no single model is universally optimal, and model
selection should be context-specific.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Despite significant investment in data collection technologies and
  logistics management systems, these systems primarily collect
  transactional data, such as consumption. However, no data is collected
  on events that influence consumption variability---information that
  would be crucial for building reliable forecasting models. No studies
  to date have explored how domain expertise could enhance
  pharmaceutical demand forecasting.
\item
  Current research on forecasting for pharmaceutical supply chain
  predominantly focuses on generating point forecasts. There is a lack
  of studies that consider the entire forecast distribution of monthly
  consumption, which would better capture the uncertainty of future
  demand and provide a valuable risk management tool for
  decision-makers.
\item
  Reproducibility remains a major challenge in this field. It is often
  difficult for readers to reproduce previous studies without direct
  assistance from the authors, limiting the practical application and
  validation of existing research.
\end{enumerate}

@@

\resizebox{\linewidth}{!}{
\begin{tabular}{lll>{\raggedright\arraybackslash}p{10em}ll>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{2em}}
\toprule
References & data granularity & Lenth of data & Forecast variable & Horizon & Probabilistic & Method & Metric & Items & Data & Code\\
\midrule
Current study & Monthly & 60 months & Consumption of pharmaceutical products & 6 months & Yes & TimeGpt (with and without predictors), LSTM (with and without predictors), Exponential Smoothing, ARIMA, Dynamic Regression (ARIMAX), Multiple Regression (with and without predictors) & RMSSE,CRIPS,MASE & 33 & Yes & Yes\\
Siddiqui et al. (2022) & Monthly & 58 months & Sales of pharmaceutical products & Un known & No & ARIMA, HOLT WINTER, ETS,  Theta, ARHOW & MEAN, MAD, MSE, RMSE,  MAPE & 31 & No & No\\
de Oliveira et al.(2021) & Daily & 365 days & Lead time & NA & No & k-NN, SVM, RF, LR, MLP & MSE & 11 & Yes & No\\
Rathipriya et al. (2023) & Weekly & 260 weeks & Sales of pharmaceuticals product & Un known & No & SNN, ARIMA P\_NN GR\_NN RBF\_NN ARIMA P\_NN GR\_NN RBF\_NN & RMSE, Normalized RMSE & 57 & Yes & No\\
Kim et al.(2015) & Monthly & 44 month & Medicines sales data & 1 month & No & VARX & Prediction error rate & 4 & No & No\\
Merkuryeva et al. (2019) & Weekly & 20 weeks & Sales of pharmaceuticals & 1–week & No & MA, Symbolic Regression, LR & R², MAD & 1 & No & No\\
Nikolopoulos et al. (2016) & Yearly & 21 year & Prescription records of pharmaceuticals & 1–5 & No & Diffusion models,  ARIMA, ES, LR & R², ME, MAE, MSE & 14 & No & No\\
Van Belle et al.(2021) & Weekly & 253 weeks & Medcine sales & 1–5 weeks & No & ARIMA, ETS, MLP, SVR, RF, LR & RMSE & 50 & No & No\\
Khalil Zadeh et al. (2014) & Monthly & 36 months & Sales  pharmaceutical products & 1 month & No & ARIMA, Graph-based analysis and ANN, & R2,MSE, MAE & 217 & No & No\\
Zhu  X et al.(2021) & Weekly & 520 weeks & Medcine sales & 1–2 months & No & MA, LR, Clustering and RNN & NME, NMAE, NMSE & 245 & No & No\\
Anusha et al.(2014) & Monthly & 36 months & Sales of pharmaceuticals & Un known & No & 6 Month MA, SES, WES & MAD, MSE, MAPE & 2 & Yes & No\\
Burinskiene (2022) & Weekly & 13 weeks & Medcine sales & 36 days & No & SES, MA, Naïve, Holt & MAPE,MSE, U² & 8 & No & No\\
Candan et al. (2014) & Yearly & 6 years & Medcine sales & 3 months & No & NF & SE Mean & 1 & No & No\\
Ribeiro et al.(2017) & Monthly & 24 months & Medcine sales & 3 months & No & DPM & SMAPE & 357 & No & No\\
Bon and NG (2017) & Monthly & 68 months & Medcine sales & 1 month & No & SMA, SES, DMA, DES, RA, HW-A, SA, HW-M, ARIMA-SM & RMSE, MSE, MAD,  MAPE  & 1 & No & No\\
Newberne and Mike (2006)  & Monthly & 36 months & Pseudoephedrine prescriptions & 3 months & No & HW-M & MAPE & 1 & No & No\\
Sousa et al.(2019) & Monthly & 79 months & drug distribution & quarterly & No & Naïve, Seasonal Naïve, ARIMA, DLM, LST, Recurrent Neural Network, SES, Holts linear and damped trends, HW, ES-SSM, Theta & MAE & 30 & Yes & Yes\\
Zahra and Putra (2019) & Monthly & 36 months & Consumption & 12 months & No & ARIMA, ES & RMSE, MAPE, MAD, MSE & 1 & No & No\\
Nguyen et  al. (2023) & Un known & Un known & Pharmaceutical products & Un known & No & VARX & ME, MAE,MAPE,RMSE & Unknown & No & No\\
Christos et al.(2017) & Weekly & 129 weeks & Web-scraping  of analgesic medcines & Un known & Yes & VARX  & AIC, BIC, MSE, MAE & Unknown & No & No\\
Belghith et al.(2024) & Daily & 1095 days & Sales of pharmaceuticals & 1 year & No & Brown’s method, Holt’s method, and Holt-Winters method, moving average & MAD, MAPE & 125 & No & No\\
Konstantinos  et al.(2024) & Weekly & 260 weeks & Medcine sales & 1 year & No & Naïve, ARIMA, XGBoost , LSTM, Prophet & MAPE,MSE & 57 & No & No\\
\bottomrule
\end{tabular}}
\endgroup{}

@@

For this study, we use a dataset spanning five years (Dec.~2017- July
2022) of consumption data obtained from EPSS. Rigorous checks were
conducted to ensure the consistency and completeness of the collected
data. From the extensive pool of pharmaceutical products within EPSS, we
selected a set of 33 key pharmaceuticals, including various programs and
representing different classes of drugs.

\subsubsection{Explaoratory Data Analysis with consumption
data}\label{explaoratory-data-analysis-with-consumption-data}

Given the number of products in this study, we created several data plot
and computed features of the time series, including the strength of
trend and seasonality to better understadn data.
Figure~\ref{fig-feature} shows the strength of trend versus seasonality.
Each point represents one time series with the strength of trend in
x-axis and the strength of seasonality in y-axis. Both measures are on a
scale of {[}0,1{]}. the strength of trend and seasonality were
calculated using the ``STL'' (Seasonal and Trend decomposition using
Loess) decomposition method, as described by \citet{bandara2025mstl}

@@

\caption{\label{fig-feature}The strength of the trend and seasonality in
the time series of pharmaceutical product consumption. The scatter plot
shows 33 data points, with each point corresponding to a product.}

@@

It is evident that some time series display strong trends and/or
seasonality, while the majority exhibit low trends and minimal
seasonality. A number of products show pronounced trends, and only a few
demonstrate clear seasonal patterns. Beyond assessing the strength of
trends and seasonality, we also visualized all time series to understand
data and various patterns, including trends and instances of erratic
consumption behavior during certain months. For example, some series
show low consumption volumes over consecutive months, followed by peak
consumption in specific months, making them more volatile and difficult
to forecast. This underscores the diversity of monthly pharmaceutical
time series patterns within the dataset and highlights the importance of
understanding the factors driving these consumption behaviors. Figure
Figure~\ref{fig-dataviz2} illustrates the time plots for a few selected
products.

@@

\caption{\label{fig-dataviz2}Monthly time plot of consumption. X-axis
shows the month, consisting of 60 data points (months) and y-axis shows
the comsumption. The panels display data from four products to give a
glimpse of the consumption patterns.}

@@

Table~\ref{tbl-feature} also provides the summary statistics computed
for each pharmaceutical product time series include traditional
distributional measures---mean, median, standard deviation, minimum, and
maximum---alongside time series--specific and structural
characteristics. These include the first-order autocorrelation (ACF at
lag 1), the mean length of consecutive zero runs (zero\_run\_mean), and
the squared coefficient of variation computed only on non-zero values
(nonzero\_squared\_cv). Together, these metrics capture not only central
tendency and dispersion, but also temporal dependence, sparsity
patterns, and relative variability in periods with non-zero consumption.

\subsection{Collaborative expert review to identify factors affecting
consumption}\label{collaborative-expert-review-to-identify-factors-affecting-consumption}

\textcolor{blue}{To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation. In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.}

Figure \ref{fig:expert-pipeline}
\textcolor{blue}{presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The missing puzzle piece: contextual insights for enhanced pharmaceutical supply chain forecasting},
  pdfauthor={Arebu Issa Bilal; Bahman Rostami-Tabar; Harsha Halgamuwe Hewage; Umit Sezer Bititci; Teferi Gedif Fenta},
  pdfkeywords={Forecasting, Pharmaceutical supply chain, Domain
knowlege, Forecast accuracy, Developing countries},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{The missing puzzle piece: contextual insights for enhanced
pharmaceutical supply chain forecasting}
\author[1]{Arebu Issa Bilal%
\corref{cor1}%
}
 \ead{arebu.issa@aau.edu.et} 
\author[2]{Bahman Rostami-Tabar%
%
}
 \ead{rostami-tabarb@cardiff.ac.uk} 
\author[2]{Harsha Halgamuwe Hewage%
%
}
 \ead{HalgamuweHewageHR@cardiff.ac.uk} 
\author[3]{Umit Sezer Bititci%
%
}
 \ead{U.S.Bititci@hw.ac.uk} 
\author[1]{Teferi Gedif Fenta%
%
}
 \ead{tgedif@gmail.com}

\affiliation[1]{organization={Addis Ababa University, Department of
Social and Administrative Pharmacy, School of
Pharmacy},country={Ethiopia.},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Cardiff University, Data Lab for Social
Good Research Group, Cardiff Business
School},country={UK.},countrysep={,},postcodesep={}}
\affiliation[3]{organization={Heriot-Watt University, Edinburgh Business
School, Heriot-Watt
University},country={UK.},countrysep={,},postcodesep={}}

@@

\begin{abstract}
Accurate forecasting in pharmaceutical supply chains is critical for
ensuring the continuous availability of essential medicines,
particularly in developing countries where resource constraints and
logistical challenges are more pronounced. Effective demand forecasting
supports procurement and inventory management, which in turn could help
to prevent stockouts, reduce wastage due to overstocking, and enhance
overall healthcare delivery. In such settings, reliable forecasts that
acknowledge uncertainties, are essential to ensure that limited
resources are used efficiently to meet health needs. Despite
advancements in forecasting, significant gaps remain in effective
forecasting in poor resource countries. Most existing literature on
pharmaceutical supply chain forecasting focuses primarily on point
estimation while neglecting the inherent uncertainty of these forecasts.
Additionally, these studies often rely solely on historical consumption
and distribution data, overlooking the broader contextual factors that
shape these patterns. Furthermore, many works lack rigorous
methodological design, transparency, and reproducibility. This study
addresses these gaps by integrating domain-specific knowledge, gathered
through expert interviews and engagement with key members of the
Ethiopian Pharmaceutical Supply Service, into forecasting models. Using
a dataset spanning five years (December 2017 to July 2022) from EPSS, we
developed forecasting models that integrate expert-identified variables
such as stock replenishment schedules, fiscal inventory counts, and
disease outbreaks. Evaluation metrics including Mean Absolute Scaled
Error, Root Mean Squared Scaled Error, and Continuous Ranked Probability
Score are used to report forecast accuracy. The findings underscore the
significance of contextual data in developing robust forecasting models
that are adaptable to complex, real-world conditions. Our results also
highlight the effectiveness of foundational time series models for
forecasting. These models are particularly appealing for
resource-constrained countries that may lack advanced analytical
expertise. To promote usability, generalizability, and reproducibility,
we share the complete dataset and code in R and Python, and the entire
paper is written in Quarto via a GitHub repository to facilitate these
practices.
\end{abstract}

@@

Demand forecasting in pharmaceutical supply chains is a highly complex
process, shaped by several factors. These include the quality of
available data, the specific position within the supply chain, and
external influences such as market trends, regulatory changes, and
disruptive events such as pandemics, conflicts, flooding
\citep{schneider2010pharmaceutical, hyndman2018forecasting}. Recently,
major donors such as Bill \& Melinda Gates Foundation have invested in
data collection technologies in sub-Saharan Africa, recognizing the
potential of data to provide valuable insights. This investment is
essential and offers significant benefits for improving supply chain
delivery. However, much of the data being collected focuses on
transactional data, such as consumption, distribution, and sale. Almost
all platforms collecting data tend to overlook the critical contextual
factors and events that influence these transactions---factors that are
just as crucial for accurate forecasting. This presents a major barrier,
as understanding the broader context, including administration
procedures, local policies, disruptions, conflicts, and so on is vital
for building reliable forecasting models. This gap --- known as business
context and domain knowledge --- requires urgent attention in
pharmaceutical supply chain forecasting. Without incorporating this
knowledge, any forecasting model is unlikely to produce reliable
results. We must prioritize collecting and integrating this contextual
information to improve the accuracy of demand forecasts in
pharmaceutical supply chains. Integrating statistical models with domain
knowledge leads to a more accurate and context-aware forecasting
process, as evidenced by research that highlights the value of
judgmental adjustments to enhance forecast accuracy
\citep{fildes2007against, taylor2022artificial, soyiri2013overview}.
Domain knowledge enables analysts to incorporate relevant events, such
as supply chain disruptions, into the forecasting model. This
integration is particularly beneficial in fields like healthcare, where
expert insights can align algorithms with real-world applications
\citep{dash2022review}.

@@

Accurate demand forecasting is crucial in the pharmaceutical industry,
where it directly impacts profit maximization, cost minimization, and
the ability to respond to market changes. Forecasting in healthcare not
only influences clinical decisions but also plays a pivotal role in
supply chain management, ensuring that the right drugs are available
when needed. The challenge lies in balancing customer demand with
inventory costs, which is particularly complex in pharmaceuticals due to
factors like short shelf life and quality constraints
\citep{gupta2000mid, makridakis2020forecasting}.

A range of methods is used in pharmaceutical supply chain forecasting,
from traditional statistical models to advanced machine learning
techniques, with many studies combining both simple and advanced
approaches to improve accuracy
\citep{nikolopoulos2016forecasting, zhu2021demand, anusha2014demand}.
Some studies, such as those by \citet{newberne2006holt}, and
\citet{restyana2021analysis}, applied traditional forecasting methods.
In contrast, other researchers explored more advanced approaches,
including hybrid models and machine learning techniques, to address
complex forecasting challenges, as seen in the studies by
\citet{siddiqui2022hybrid}, \citet{de2021lead}, \citet{kim2015demand},
\citet{candan2014demand}, and \citet{ribeiro2017improving}.

Early research in pharmaceutical forecasting primarily focused on
classical statistical methods. These methods, including ARIMA
(AutoRegressive Integrated Moving Average), exponential smoothing, and
moving averages, were widely used due to their simplicity and
effectiveness in relatively stable environments
\citep{zahra2019forecasting}. These models typically assess forecast
accuracy using metrics like Mean Absolute Percentage Error (MAPE), Mean
Squared Error (MSE), and Root Mean Squared Error (RMSE). For example,
\citet{newberne2006holt} applied the Holt-Winters method to forecast
healthcare data, focusing on prescription trends. Their findings
highlighted the method's utility in short-term planning and resource
management, validated through metrics like MAPE, MAD, and MSD.
Similarly, \citet{restyana2021analysis} compared Simple Moving Average
(SMA), and Single Exponential Smoothing (SES) in drug demand
forecasting, concluding that SES provided more accurate results, as
indicated by lower MAD and MSE values.

Recognizing the limitations of traditional models---particularly their
inability to capture the complexities of pharmaceutical demand, such as
seasonality and external influences---researchers have increasingly
turned to more advanced modeling approaches
\citep{khalil2014intelligent}. Furthermore, the growing complexity of
the pharmaceutical industry, coupled with the availability of large
datasets, has accelerated the adoption of advanced machine learning
techniques. These methods, such as Support Vector Regression (SVR),
Random Forest (RF), and Long Short-Term Memory (LSTM) networks, offer
greater flexibility and predictive power. \citet{van2021using} showed
that incorporating downstream information into machine learning models,
like LASSO and SVR, greatly enhances forecast accuracy. Their study on
multi-echelon supply chains revealed that advanced models, particularly
those integrating external variables, consistently produced lower
forecast errors, as measured by AvgRelRMSE. Similarly,
\citet{rathipriya2023demand} compared shallow neural networks and deep
learning models for drug demand forecasting. The study found that
shallow models outperformed deep learning models for most drug
categories, while ARIMA was more effective for the remaining categories.
This suggests that no single model is universally optimal, and model
selection should be context-specific.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Despite significant investment in data collection technologies and
  logistics management systems, these systems primarily collect
  transactional data, such as consumption and distribution. However, no
  data is collected on events that influence their variability ---
  information that would be crucial for building reliable forecasting
  models. No studies to date have explored how domain expertise could
  enhance pharmaceutical demand forecasting.
\item
  Current research on forecasting for pharmaceutical supply chain
  predominantly focuses on generating point forecasts. There is a lack
  of studies that consider the entire forecast distribution, which would
  better capture the uncertainty of future demand and provide a valuable
  risk management tool for decision-makers.
\item
  Reproducibility remains a major challenge in this field. It is often
  difficult for readers to reproduce previous studies without direct
  assistance from the authors, limiting the practical application and
  validation of existing research.
\end{enumerate}

@@

\resizebox{\linewidth}{!}{
\begin{tabular}{lll>{\raggedright\arraybackslash}p{10em}ll>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{2em}}
\toprule
References & data granularity & Lenth of data & Forecast variable & Horizon & Probabilistic & Method & Metric & Items & Data & Code\\
\midrule
Current study & Monthly & 60 months & Consumption of pharmaceutical products & 6 months & Yes & TimeGpt (with and without predictors), LSTM (with and without predictors), Exponential Smoothing, ARIMA, Dynamic Regression (ARIMAX), Multiple Regression (with and without predictors) & RMSSE,CRIPS,MASE & 33 & Yes & Yes\\
Siddiqui et al. (2022) & Monthly & 58 months & Sales of pharmaceutical products & Unknown & No & ARIMA, HOLT WINTER, ETS,  Theta, ARHOW & MEAN, MAD, MSE, RMSE,  MAPE & 31 & No & No\\
de Oliveira et al.(2021) & Daily & 365 days & Lead time & NA & No & k-NN, SVM, RF, LR, MLP & MSE & 11 & Yes & No\\
Rathipriya et al. (2023) & Weekly & 260 weeks & Sales of pharmaceuticals product & Unknown & No & SNN, ARIMA P\_NN GR\_NN RBF\_NN ARIMA P\_NN GR\_NN RBF\_NN & RMSE, Normalized RMSE & 57 & Yes & No\\
Kim et al.(2015) & Monthly & 44 month & Medicines sales data & 1 month & No & VARX & Prediction error rate & 4 & No & No\\
Merkuryeva et al. (2019) & Weekly & 20 weeks & Sales of pharmaceuticals & 1–week & No & MA, Symbolic Regression, LR & R², MAD & 1 & No & No\\
Nikolopoulos et al. (2016) & Yearly & 21 year & Prescription records of pharmaceuticals & 1–5 & No & Diffusion models,  ARIMA, ES, LR & R², ME, MAE, MSE & 14 & No & No\\
Van Belle et al.(2021) & Weekly & 253 weeks & Medcine sales & 1–5 weeks & No & ARIMA, ETS, MLP, SVR, RF, LR & RMSE & 50 & No & No\\
Khalil Zadeh et al. (2014) & Monthly & 36 months & Sales  pharmaceutical products & 1 month & No & ARIMA, Graph-based analysis and ANN, & R2,MSE, MAE & 217 & No & No\\
Zhu  X et al.(2021) & Weekly & 520 weeks & Medcine sales & 1–2 months & No & MA, LR, Clustering and RNN & NME, NMAE, NMSE & 245 & No & No\\
Anusha et al.(2014) & Monthly & 36 months & Sales of pharmaceuticals & Un known & No & 6 Month MA, SES, WES & MAD, MSE, MAPE & 2 & Yes & No\\
Burinskiene (2022) & Weekly & 13 weeks & Medcine sales & 36 days & No & SES, MA, Naïve, Holt & MAPE,MSE, U² & 8 & No & No\\
Candan et al. (2014) & Yearly & 6 years & Medcine sales & 3 months & No & NF & SE Mean & 1 & No & No\\
Ribeiro et al.(2017) & Monthly & 24 months & Medcine sales & 3 months & No & DPM & SMAPE & 357 & No & No\\
Bon and NG (2017) & Monthly & 68 months & Medcine sales & 1 month & No & SMA, SES, DMA, DES, RA, HW-A, SA, HW-M, ARIMA-SM & RMSE, MSE, MAD,  MAPE  & 1 & No & No\\
Newberne and Mike (2006)  & Monthly & 36 months & Pseudoephedrine prescriptions & 3 months & No & HW-M & MAPE & 1 & No & No\\
Sousa et al.(2019) & Monthly & 79 months & drug distribution & quarterly & No & Naïve, Seasonal Naïve, ARIMA, DLM, LST, Recurrent Neural Network, SES, Holts linear and damped trends, HW, ES-SSM, Theta & MAE & 30 & Yes & Yes\\
Zahra and Putra (2019) & Monthly & 36 months & Consumption & 12 months & No & ARIMA, ES & RMSE, MAPE, MAD, MSE & 1 & No & No\\
Nguyen et  al. (2023) & Un known & Un known & Pharmaceutical products & Unknown & No & VARX & ME, MAE,MAPE,RMSE & Unknown & No & No\\
Christos et al.(2017) & Weekly & 129 weeks & Web-scraping  of analgesic medcines & Unknown & Yes & VARX  & AIC, BIC, MSE, MAE & Unknown & No & No\\
Belghith et al.(2024) & Daily & 1095 days & Sales of pharmaceuticals & 1 year & No & Brown’s method, Holt’s method, and Holt-Winters method, moving average & MAD, MAPE & 125 & No & No\\
Konstantinos  et al.(2024) & Weekly & 260 weeks & Medcine sales & 1 year & No & Naïve, ARIMA, XGBoost , LSTM, Prophet & MAPE,MSE & 57 & No & No\\
\bottomrule
\end{tabular}}
\endgroup{}

@@

For this study, we use a dataset spanning five years (December 2017 -
July 2022) of distribution data obtained from EPSS. Rigorous checks were
conducted to ensure the consistency and completeness of the collected
data. From the extensive pool of pharmaceutical products within EPSS, we
selected a set of 33 key pharmaceuticals, including various programs and
representing different classes of drugs.

\subsubsection{Exploratory Data Analysis with distribution
data}\label{exploratory-data-analysis-with-distribution-data}

Given the number of products in this study, we created several data plot
and computed features of the time series, including the strength of
trend and seasonality to better understand data.
Figure~\ref{fig-feature} shows the strength of trend versus seasonality.
Each point represents one time series with the strength of trend in
x-axis and the strength of seasonality in y-axis. Both measures are on a
scale of {[}0,1{]}. the strength of trend and seasonality were
calculated using the ``STL'' (Seasonal and Trend decomposition using
Loess) decomposition method, as described by \citet{bandara2025mstl}

@@

\caption{\label{fig-feature}The strength of the trend and seasonality in
the time series of pharmaceutical product distribution. The scatter plot
shows 33 data points, with each point corresponding to a product.}

@@

It is evident that some time series display strong trends and/or
seasonality, while the majority exhibit low trends and minimal
seasonality. A number of products show pronounced trends, and only a few
demonstrate clear seasonal patterns. Beyond assessing the strength of
trends and seasonality, we also visualized all time series to understand
data and various patterns, including trends and instances of erratic
distribution behavior during certain months. For example, some series
show low distribution volumes over consecutive months, followed by peak
distribution in specific months, making them more volatile and difficult
to forecast. This underscores the diversity of monthly pharmaceutical
time series patterns within the dataset and highlights the importance of
understanding the factors driving these distribution behaviors.
Figure~\ref{fig-dataviz2} illustrates the time plots for a few selected
products.

@@

\caption{\label{fig-dataviz2}Monthly time plot of distribution. X-axis
shows the month, consisting of 60 data points (months) and y-axis shows
the comsumption. The panels display data from four products to give a
glimpse of the distribution patterns.}

@@

Table~\ref{tbl-feature} also provides the summary statistics computed
for each pharmaceutical product time series include traditional
distributional measures---mean, median, standard deviation, minimum, and
maximum---alongside time series--specific and structural
characteristics. These include the first-order autocorrelation (ACF at
lag 1), the mean length of consecutive zero runs (zero\_run\_mean), and
the squared coefficient of variation computed only on non-zero values
(nonzero\_squared\_cv). Together, these metrics capture not only central
tendency and dispersion, but also temporal dependence, sparsity
patterns, and relative variability in periods with non-zero
distribution.

\subsection{Collaborative expert review to identify factors affecting
distribution}\label{collaborative-expert-review-to-identify-factors-affecting-distribution}

To incorporate expert knowledge into time series forecasting in a
structured and replicable manner, we developed a multi-step process
involving expert engagement, collaborative data review, and operational
validation. In partnership with the Forecasting and Market Shaping
Directorate at EPSS, we identified six experienced professionals from
the Warehouse and Inventory Management Directorate with deep operational
knowledge of pharmaceutical logistics.

Figure \ref{fig:expert-pipeline} presents a visual overview of the
end-to-end approach, highlighting each major stage of the process from
expert engagement to model integration.
